---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Floodplain Delineation {-#app-floodplain}

```{r flood-setup, include = FALSE}
library(terra)
library(sf)
```

```{r flood-load}
gpkg    <- "data/gis/necr.gpkg"
dem_tif <- "data/gis/necr_dem.tif"
val_tif <- "data/gis/necr_valleys.tif"

present <- sf::st_layers(gpkg)$name
load_layer <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi           <- load_layer("aoi")
streams       <- load_layer("streams")
waterbodies   <- load_layer("waterbodies")
floodplain    <- load_layer("floodplain")
railways      <- load_layer("railways")
roads         <- load_layer("roads")
reserves      <- load_layer("reserves")
parks         <- load_layer("parks")
named_streams <- load_layer("named_streams")
dem     <- terra::rast(dem_tif)
valleys <- terra::rast(val_tif)

dem        <- terra::mask(terra::crop(dem,     terra::vect(aoi)), terra::vect(aoi))
valleys    <- terra::mask(terra::crop(valleys, terra::vect(aoi)), terra::vect(aoi))
streams     <- suppressWarnings(sf::st_intersection(streams,     aoi))
waterbodies <- suppressWarnings(sf::st_intersection(waterbodies, aoi))
floodplain  <- suppressWarnings(sf::st_intersection(floodplain,  aoi))
if (!is.null(roads))         roads         <- suppressWarnings(sf::st_intersection(roads,         aoi))
if (!is.null(railways))      railways      <- suppressWarnings(sf::st_intersection(railways,      aoi))
if (!is.null(reserves))      reserves      <- suppressWarnings(sf::st_intersection(reserves,      aoi))
if (!is.null(parks))         parks         <- suppressWarnings(sf::st_intersection(parks,         aoi))
if (!is.null(named_streams)) named_streams <- suppressWarnings(sf::st_intersection(named_streams, aoi))
```

```{r flood-rollup}
aoi_area_km2 <- as.numeric(sum(sf::st_area(aoi))) / 1e6
fp_area_ha   <- as.numeric(sum(sf::st_area(floodplain))) / 1e4
fp_pct_aoi   <- 100 * fp_area_ha / (aoi_area_km2 * 100)

streams_in_fp    <- suppressWarnings(sf::st_intersection(streams, floodplain))
streams_km_total <- as.numeric(sum(sf::st_length(streams)))    / 1e3
streams_km_in_fp <- as.numeric(sum(sf::st_length(streams_in_fp))) / 1e3

wb_in_fp    <- suppressWarnings(sf::st_intersection(waterbodies, floodplain))
lakes_n     <- sum(wb_in_fp$waterbody_type == "L", na.rm = TRUE)
wetlands_n  <- sum(wb_in_fp$waterbody_type == "W", na.rm = TRUE)
lakes_ha    <- as.numeric(sum(sf::st_area(wb_in_fp[wb_in_fp$waterbody_type == "L", ]))) / 1e4
wetlands_ha <- as.numeric(sum(sf::st_area(wb_in_fp[wb_in_fp$waterbody_type == "W", ]))) / 1e4
```

## Floodplains and fish passage {-}

Most fish-passage assessments treat a stream crossing as a single point
on the mainstem — does it pass flow at bankfull, what is the gradient
through the structure, can adult and juvenile fish move through it.
That view captures only half the story. Streams in this region run
through floodplains: the lower-lying ground that water reclaims at high
flows, threaded with side channels, sloughs, beaver complexes, ponds,
and wetlands. For salmonids, those off-channel areas are where freshwater survival
is often won or lost. Winter low-flow refuge, freshet high-flow refuge,
and the slow-water rearing that juveniles depend on are floodplain
features, not mainstem features.

A barrier on the mainstem that severs lateral connection to its
floodplain therefore cuts fish off from more than the river upstream —
it cuts them off from the off-channel network that gives the reach its
productivity. Mapping floodplain extent at watershed scale lets us
overlay this information on the linear barrier inventory and ask a
second question alongside passage: how much off-channel habitat sits
within the affected footprint, and what does it contain.

We piloted floodplain delineation for the Nechako River Watershed Group
this reporting year. In coming years we plan to extend the method to
the remaining watersheds so that floodplain extent and
values can carry through to barrier prioritisation across the region.

## Methods {-}

Floodplain footprints were delineated with the
[`flooded`](https://newgraphenvironment.github.io/flooded/) R package,
which combines a digital elevation model, the stream network, and a
modelled bankfull flood surface to identify the low-relief ground that
water can reach at high flow. The underlying algorithm is the Valley
Confinement Algorithm of @nagel_etal2014LandscapeScale, with bankfull
depth from the regional regression of @hall_etal2007Predictingriver.

For this pilot we used the **functional floodplain** scenario
(`ff04` in `flooded::fl_scenarios()`) — the recurrent-inundation
footprint produced by the bankfull-depth regression
@hall_etal2007Predictingriver calibrated against field-mapped
floodplain sites in Pacific Northwest streams. Higher and lower
flood-factor settings trace the active channel margin or the broader
valley bottom; the functional setting was chosen as the most relevant
scale for fish-habitat planning.

Inputs were the
[MRDEM-30](https://open.canada.ca/data/en/dataset/18752265-bda3-498c-a4ba-9dfe68cb98da)
30 m digital elevation model, the chinook accessible stream network
(stream order ≥ 3) from
[bcfishpass](https://github.com/smnorris/bcfishpass), and BC Freshwater
Atlas lakes and wetlands attached to that network.

## Results {-}

```{r flood-table}
knitr::kable(
  data.frame(
    Metric = c(
      "Watershed group area (km²)",
      "Floodplain area (ha)",
      "Floodplain as % of watershed group",
      "Chinook accessible streams, order 3+ (km)",
      "Streams within floodplain (km)",
      "Lakes within floodplain (count)",
      "Lakes within floodplain (ha)",
      "Wetlands within floodplain (count)",
      "Wetlands within floodplain (ha)"
    ),
    Value = c(
      format(round(aoi_area_km2),     big.mark = ","),
      format(round(fp_area_ha),       big.mark = ","),
      sprintf("%.1f", fp_pct_aoi),
      format(round(streams_km_total), big.mark = ","),
      format(round(streams_km_in_fp), big.mark = ","),
      format(lakes_n,                 big.mark = ","),
      format(round(lakes_ha),         big.mark = ","),
      format(wetlands_n,              big.mark = ","),
      format(round(wetlands_ha),      big.mark = ",")
    ),
    stringsAsFactors = FALSE
  ),
  caption = "Nechako River Watershed Group — modelled functional floodplain and the off-channel habitat units within it.",
  col.names = c("Metric", "Value"),
  align = c("l", "r")
)
```

The Nechako River Watershed Group covers roughly
`r format(round(aoi_area_km2), big.mark = ",")` km². Modelled
functional floodplain
(Table \@ref(tab:flood-table)) covers about
`r format(round(fp_area_ha), big.mark = ",")` ha, or
`r sprintf("%.1f", fp_pct_aoi)` % of the watershed group, attached to
`r format(round(streams_km_in_fp), big.mark = ",")` km of the
`r format(round(streams_km_total), big.mark = ",")` km chinook
accessible order-3+ network. Within that footprint sit `r lakes_n`
lakes and `r format(wetlands_n, big.mark = ",")` mapped wetland
polygons — the off-channel habitat units that lateral floodplain
connectivity supports.

The watershed-wide footprint is shown in
Figure \@ref(fig:flood-map-wsg); a detail view of the Murray Creek
confluence area is shown in Figure \@ref(fig:flood-map-detail).

```{r flood-map-wsg, fig.cap = "Nechako River Watershed Group functional floodplain (green) over hillshaded terrain. Chinook accessible order-3+ streams in blue; lakes and wetlands in light blue; parks light green; First Nations reserves light grey with diamond markers and labels; forest service / resource roads grey; railways black dashed; watershed boundary heavy black."}
hill <- terra::shade(
  terra::terrain(dem, "slope",  unit = "radians"),
  terra::terrain(dem, "aspect", unit = "radians")
)

terra::plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 4, 2),
            main = "Nechako River Watershed Group\nfunctional floodplain, accessible order 3+")

terra::plot(valleys, col = c(NA, "#2c7a2c"), alpha = 0.55, add = TRUE, legend = FALSE)
plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
     lwd = 0.3, add = TRUE)
plot(sf::st_geometry(streams), col = "#1f77b4", lwd = 0.4, add = TRUE)

if (!is.null(roads)) {
  fsr <- roads[roads$transport_line_type_code %in% c("RRS", "RRD", "RRN"), ]
  if (nrow(fsr) > 0L) {
    plot(sf::st_geometry(fsr), col = "#88888888", lwd = 0.25, add = TRUE)
  }
}
if (!is.null(railways)) {
  plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed", add = TRUE)
}
if (!is.null(parks)) {
  plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
       lwd = 0.7, add = TRUE)
}
if (!is.null(reserves)) {
  plot(sf::st_geometry(reserves), col = "#b2b2b288", border = "#232323",
       lwd = 0.6, add = TRUE)
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 0.9, add = TRUE)
}

plot(sf::st_geometry(aoi), border = "black", lwd = 1.5, add = TRUE)

if (!is.null(reserves) && "english_name" %in% names(reserves) &&
    nrow(reserves) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) text(coords[, 1] + dx * halo_x, ly,
                            labels = reserves$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  for (dy in c(-1, 1)) text(coords[, 1], ly + dy * halo_y,
                            labels = reserves$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  text(coords[, 1], ly, labels = reserves$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

```{r flood-map-detail, fig.cap = "Detail view of the Murray Creek confluence with the Nechako River. Symbology as Figure \\@ref(fig:flood-map-wsg); named streams labelled in italic blue. The town of Vanderhoof is visible at the bottom right."}
# Murray Creek confluence with Nechako River — Vanderhoof at bottom right.
# We locate Murray Creek from named_streams, then build a bbox that captures
# the upstream section of Murray Creek, the confluence, and extends south-east
# to include Vanderhoof.
murray <- named_streams[grepl("Murray", named_streams$gnis_name, ignore.case = TRUE), ]
if (nrow(murray) > 0L) {
  mb <- sf::st_bbox(murray)
  # Extend south-east to capture Vanderhoof
  inset_bbox <- sf::st_bbox(c(
    xmin = unname(mb["xmin"] - (mb["xmax"] - mb["xmin"]) * 0.15),
    ymin = unname(mb["ymin"] - (mb["ymax"] - mb["ymin"]) * 0.6),
    xmax = unname(mb["xmax"] + (mb["xmax"] - mb["xmin"]) * 0.5),
    ymax = unname(mb["ymax"] + (mb["ymax"] - mb["ymin"]) * 0.15)
  ), crs = sf::st_crs(streams))
} else {
  # Fallback: centre of WSG
  e <- sf::st_bbox(aoi)
  cx <- (e["xmin"] + e["xmax"]) / 2
  cy <- (e["ymin"] + e["ymax"]) / 2
  span_x <- (e["xmax"] - e["xmin"]) * 0.3
  span_y <- (e["ymax"] - e["ymin"]) * 0.3
  inset_bbox <- sf::st_bbox(c(
    xmin = unname(cx - span_x), ymin = unname(cy - span_y),
    xmax = unname(cx + span_x), ymax = unname(cy + span_y)
  ), crs = sf::st_crs(streams))
}
inset_ext <- terra::ext(inset_bbox[c("xmin", "xmax", "ymin", "ymax")])

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

dem_inset           <- terra::crop(dem, inset_ext)
valleys_inset       <- terra::crop(valleys, inset_ext)
streams_inset       <- crop_sf(streams)
wb_inset            <- crop_sf(waterbodies)
aoi_inset           <- crop_sf(aoi)
roads_inset         <- crop_sf(roads)
railways_inset      <- crop_sf(railways)
reserves_inset      <- crop_sf(reserves)
parks_inset         <- crop_sf(parks)
named_streams_inset <- crop_sf(named_streams)

hill_inset <- terra::shade(
  terra::terrain(dem_inset, "slope",  unit = "radians"),
  terra::terrain(dem_inset, "aspect", unit = "radians")
)

terra::plot(hill_inset, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 4, 2),
            main = "Nechako River Watershed Group\nMurray Creek confluence detail")

terra::plot(valleys_inset, col = c(NA, "#2c7a2c"), alpha = 0.55,
            add = TRUE, legend = FALSE)
if (!is.null(wb_inset)) {
  plot(sf::st_geometry(wb_inset), col = "#a3cdb988", border = "#2171B5",
       lwd = 0.4, add = TRUE)
}
plot(sf::st_geometry(streams_inset), col = "#1f77b4", lwd = 0.7, add = TRUE)

if (!is.null(roads_inset)) {
  plot(sf::st_geometry(roads_inset), col = "#666666aa", lwd = 0.4, add = TRUE)
}
if (!is.null(railways_inset)) {
  plot(sf::st_geometry(railways_inset), col = "black", lwd = 1.0, lty = "dashed",
       add = TRUE)
}
if (!is.null(parks_inset)) {
  plot(sf::st_geometry(parks_inset), col = "#639b5f55", border = "#33a02c",
       lwd = 0.8, add = TRUE)
}
if (!is.null(reserves_inset)) {
  plot(sf::st_geometry(reserves_inset), col = "#b2b2b2aa", border = "#232323",
       lwd = 0.7, add = TRUE)
}
if (!is.null(aoi_inset)) {
  plot(sf::st_geometry(aoi_inset), border = "black", lwd = 1.5, add = TRUE)
}

if (!is.null(named_streams_inset)) {
  dedup <- named_streams_inset[!duplicated(named_streams_inset$gnis_name), ]
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
  text(sf::st_coordinates(pts), labels = dedup$gnis_name,
       cex = 0.5, font = 3, col = "#0d3a6c")
}

if (!is.null(reserves_inset) && "english_name" %in% names(reserves_inset) &&
    nrow(reserves_inset) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves_inset),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 1.0, add = TRUE)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) text(coords[, 1] + dx * halo_x, ly,
                            labels = reserves_inset$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  for (dy in c(-1, 1)) text(coords[, 1], ly + dy * halo_y,
                            labels = reserves_inset$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  text(coords[, 1], ly, labels = reserves_inset$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

The watershed-wide view shows the floodplain following the trunk
Nechako River, with substantial off-channel area in the broader valley
reaches and at tributary confluences. The detail view resolves the
per-reach floodplain pattern around the Murray Creek confluence and
gives a sense of the resolution available within any sub-area for
downstream analyses — barrier prioritisation, land-cover change
tracking, or restoration scoping.
