---
title: "0210_fiss_export_to_template"
date: "Created: 2024-06-20 | Updated: `r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  html_document:
    code_folding: "hide"
editor_options:
  chunk_output_type: console
---

# Purpose of this script

Export fiss site data to csv for cut and paste into the fish data submission spreadsheet. This script has been updated so that the csv contains as much of the data as possible, so that only a few copy/pastes are required. We are trying to move away from all the back and forths between the spreadsheet, R, and Q.

```{r params}
# Params come from the report front matter so there is one source of truth, and
# so this script satisfies the `01_prep_inputs/README.md` contract of being
# self-sufficient in a clean environment. Same pattern as `scripts/run_pagedown.R`.
params <- rmarkdown::yaml_front_matter("index.Rmd")$params

path_form_fiss_site <- fs::path_expand(fs::path('~/Projects/gis/', params$gis_project_name ,'/data_field/', params$project_year, paste0("form_fiss_site_", params$project_year, ".gpkg")))

# Blank provincial FDS template - read only, to harvest target column names.
# The populated workbook is no longer produced (see #23), so the names come from
# the blank rather than from a filled `habitat_confirmations.xls`.
path_fds_template <- "data/templates/FDS_Template2026-03-11.xlsx"
```



# Prepare the data for copy/paste into submission spreadsheet

After QAing the data in QGIS, now we will prepare it for copy/paste into submission spreadsheet.


## Backup the form after the QA

```{r backup}

# Returns the QA'd form with UTM columns refreshed from geometry. This is the
# data source for everything below - `0205_fiss_wrangle.R` writes the wrangled
# form back to the gpkg, then removes its own `form_fiss_site_prep*` objects
# (`0205:191`), so they are not available here.
form_fiss_site_qa <- fpr::fpr_sp_gpkg_backup(
  path_gpkg = path_form_fiss_site,
  dir_backup = "data/backup/",
  update_utm = TRUE,
  update_site_id = FALSE,
  write_back_to_path = FALSE,
  return_object = TRUE,
  col_easting = "utm_easting",
  col_northing = "utm_northing"
)
```


## Prepare the data for copy paste to spreadsheet.

```{r cp-prep}
# see the names of our form
names(form_fiss_site_qa)

# let's get the names of the input template
# there is lots of work to do to pull out all the information we can use so we will start with one small step at a time
# lets just populate the location and site info pages.
# Read the blank template once - it is ~3.7 MB across 8 sheets, so re-reading it
# per lookup is wasteful. Plucked by name rather than position so a sheet order
# change in a future template version fails loudly instead of silently.
fds_names <- fpr::fpr_import_hab_con(path = path_fds_template,
                                     backup = F,
                                     row_empty_remove = T)

form_raw_names_site <- fds_names |>
  purrr::pluck("step_4_stream_site_data") |>
  names()

# location names
form_raw_names_location <- fds_names |>
  purrr::pluck("step_1_ref_and_loc_info") |>
  names()



# we don't want duplicate column names because it messes with them (renames them both) so we need to get rid of dupes
names_dup <- intersect(form_raw_names_site, form_raw_names_location)

# join the names of our two target tables together without (!) the dupes
form_raw_names_sl <- c(form_raw_names_location,
                       form_raw_names_site[!form_raw_names_site %in% names_dup])


# Rename the columns to match those in the spreadsheet
form_site_info_prep <- form_fiss_site_qa |>
  sf::st_drop_geometry() |>
  # alias local name and gazetted_name is not called the same in both sheets so rename
  dplyr::rename(alias_local_name = local_name,
                gazetted_name = gazetted_names,
                waterbody_id_identifier = waterbody_id,
                watershed_code_45_digit = watershed_code_50k) |> 
  dplyr::mutate(utm_method = "GPS General",
                reference_number = dplyr::row_number()) |> 
  dplyr::select(mergin_user,
                date_time_start,
                dplyr::contains('surveyor'),
                dplyr::any_of(form_raw_names_sl))

```


## Filter and c/p the reference and location data

Filter out the location data and burn to csv for copy paste into `step_1_ref_and_loc_info`

```{r cp-ref-location}
# make the reference and location form to c.p into `step_1_ref_and_loc_info`
form_fiss_loc <- bind_rows(

  # we need the raw form or we don't have all the right columns
  fds_names |>
    purrr::pluck("step_1_ref_and_loc_info") |>
    dplyr::slice(0) |>
    dplyr::mutate(dplyr::across(dplyr::everything(), as.character)),
      
   form_site_info_prep |> 
    dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) |> 
    dplyr::select(dplyr::any_of(form_raw_names_location))) |> 
  
  dplyr::mutate(site_number = dplyr::row_number())


# burn to file
form_fiss_loc |> 
  readr::write_csv(paste0(
    'data/inputs_extracted/form_fiss_loc_tidy',
    '.csv'),
    na = '')

# Now copy paste the date from the csv into `step_1_ref_and_loc_info`
```


## Filter and c/p stream site data

Filter out the stream site data and burn to csv for copy paste into `step_4_stream_site_data`

```{r cp-stream-site}
# make the stream site form to c/p into `step_4_stream_site_data`
form_fiss_site <- bind_rows(

  # we need the raw form or we don't have all the right columns
  fds_names |>
    purrr::pluck("step_4_stream_site_data") |>
    dplyr::slice(0) |>
    dplyr::mutate(dplyr::across(dplyr::everything(), as.character)),
                   
  form_site_info_prep |> 
    dplyr:: mutate(dplyr::across(dplyr::everything(), as.character)) |>
    # these column names differ slightly from those in `step_1_ref_and_loc_info`, so we update them so they are included in the csv for easier c/p.
    dplyr::rename(local_name = alias_local_name,
                gazetted_names = gazetted_name,
                waterbody_id = waterbody_id_identifier) |> 
    select(dplyr::any_of(form_raw_names_site),
           # add the time to help put the puzzle together after)
           survey_date)
) |> 
  select(everything())


# burn to file
form_fiss_site |> 
  readr::write_csv(paste0(
    'data/inputs_extracted/form_fiss_site_tidy',
    '.csv'),
    na = '')


# Now copy paste the date from the csv into `step_4_stream_site_data`

```

Next populate step 3 in fish_data_tidy.R then populate step 2 in 0100-extract-inputs.R.


